---
title: "data_processing_notes"
output: html_document
date: "2022-07-13"
---

## Introduction to dplyr: 

dplyr is an R package in the tidyverse used for data manipulation. 
```{r} 
# loading package 
library(dplyr) 

# previewing the dataset 
head(starwars) 
#note: this gives you the first 6 rows, but if you want to see a different number of rows then you can add a number after the dataset name when calling the function 

``` 

## Filtering Data: 

We can filter data using `filter()`. This allows us to subset observations (rows) based on their values (in columns). 

```{r} 

filter(starwars, hair_color == "blond" | hair_color == "blonde")  
# gets rid of anything that doesn't match the parameters 
# note: when you want to filter something, you can filter out as many things as you want 
filter(starwars, hair_color == "blond" & species == "Human") 

# it does each one at a time -- first it filters out all the humans, and then it filters the next from that group, etc 
filter(starwars, species == "Human", homeworld == "Tatooine", skin_color == "fair") 

filteredData = filter(starwars, species == "Human", homeworld == "Tatooine", skin_color == "fair") 

# filtering out observations (rows) with variables that match certain specific values (you give the filter function a vector of possible values): 
filter(starwars, hair_color %in% c("blond", "blonde")) 

# the list of possible values can also be defined before the filter function is called: 
importantPlanets = c("Tatooine", "Naboo", "Hoth") 
filter(starwars, homeworld %in% importantPlanets) 

# including missing values: 
filter(starwars, mass != 75 | is.na(mass), name < "Mace") 

# filtering out the characters that are missing a hair color 
filter(starwars, is.na(hair_color)) 

```

#### Misc. Notes: 

* or: | 
* and: & 
* less than or equal to: <= 
* greater than or equal to: >= 
* not: ! 
* when you type "variable != something" it automatically assumes you don't want to include any missing (NA) values 
* NA: missing value; can be any type 
* for exponents, you can use either ^ or ** 
* with filter and arrange, a comma means and; with select, a comma means or 
* shortcut for when you want a value between 2 different numbers: between function, ex. between(birth_year, 100, 200) 
* the summary function works differently for different types
* if you use a < or > with a string, it'll sort it alphabetically 
* it's not really based on ASCII values, but basically letters earlier in the alphabet < letters later in the alphabet, ignoring case 
* if there's an error when using the filter function, make sure the column name is spelled correctly (and the value name if it's a categorical variable) 
* make sure to use `==` when comparing observations (NOT `=` -- that's an assignment operator) 
* <, >, <=, and >= cna be used to compare numeric or categorical variables (nominal variables - ranked alphabetically as described above, while **ordinal variables have a built-in rank**) 

```{r} 
# loading the library (the diamonds dataset is inside the ggplot2 library) 
library(ggplot2) 
# ranked data 
head(diamonds) # displays the first 6 rows 
class(diamonds$cut) # gives you the specific type of data 
summary(diamonds$cut) # gives you a count of each category, or summary statistics if numeric 
summary(diamonds$carat) 
head(diamonds$cut) # displays the first 6 values (and levels) 

filter(diamonds, cut > "Good") #shows all the diamonds that are very good, premium, or ideal 

## ordering categorical data: 
# the unique function returns all unique values in a vector 
unique(starwars$eye_color) 
# defining how you want it to be ranked: 
# note: all unique values need to be listed in the order you want them in, and if you don't list them all then you'll get some NA values 
factor(starwars$eye_color, c("red", "orange", "gold", "yellow", "green", "blue", "black"), ordered = T) 

# filtering out all the characters that are shorter than 100 cm 
filter(starwars, height < 100, species != "Droid")  

# filtering out the characters that were born between the years 100 and 200 (inclusive) 
filter(starwars, birth_year >= 100 & birth_year <= 200) 

# shortcut (using the between function): 
filter(starwars, between(birth_year, 100, 200)) 

# filtering out the characters that weight over 100kg but are shorter than 185 cm 
filter(starwars, mass > 100, height < 185) 

## Missing Values: 
NA > 5 # this doesn't cause an error, but it returns NA because it doesn't know -- there's no way to determine the answer 

# function to determine if a value is NA: 
is.na(NA) 

mass = NA 
mass == 10 # similar to the above line 
is.na(mass) # returns true 
NA + 10 
NA / 2 

mary_age = NA 
paul_age = NA 
mary_age == paul_age # still returns NA because the values are unknown 

```

## Arranging Data: 

`arrange()` reorders rows. It doesn't remove any rows. <br> 
NA values are always at the end when you order by a column. 

```{r} 

# arranging birth year from lowest to highest 
arrange(starwars, birth_year) 

# highest to lowest birth year 
arrange(starwars, desc(birth_year))  

# categorical - alphabetical order 
arrange(starwars, hair_color) 

# multiple columns: 
arrange(starwars, hair_color, birth_year) 

## practice: 

# arranging the characters by height in order to find the tallest and shortest characters 
arrange(starwars, height) 
arrange(starwars, desc(height)) 

## alphabetizing the star wars characters by name 
arrange(starwars, name) 

# using arrange() to sort all missing values to the start 

# these two do the same thing, and they're not exactly right 
arrange(starwars, desc(is.na(height))) 
arrange(starwars, !is.na(height)) 

arrange(starwars, desc(is.na(height)), desc(is.na(mass))) 

arrange(starwars, !is.na(starwars)) 

```

## Selecting Certain Columns in a Dataset: 

`select()` allows us to retain only certain variables (columns). It doesn't change the order, but it removes any columns not named. 

```{r} 

select(starwars, hair_color, skin_color, eye_color) 
# since the columns listed in the above line are all in the same order that they appear in the list, this line does the same thing: 
select(starwars, hair_color:eye_color) 

# if you want everything except a certain column: 
select(starwars, -hair_color) 

# note: when you want to see everything except for several columns, you need parentheses: 
select(starwars, -(hair_color:eye_color)) 

starwars_no_color = select(starwars, -(hair_color:eye_color)) 

# filtering out all columns that contain a certain string in the title 
select(starwars, contains("color")) 

select(starwars, contains("_")) 

select(starwars, ends_with("color")) 

select(starwars, starts_with("s")) 

# the comma is kinda like an or -- if you want everything to start with s AND end with color then you have to use & 
select(starwars, starts_with("s"), ends_with("color")) 

# renaming columns: 
rename(starwars, birthYear = birth_year) 

## practice 

# select out the homeworld and species for the starwars dataset. What question might this subset of variables help answer? 
select(starwars, homeworld, species) 

# brainstorm as many ways possible to select hair_color, eye_color, skin_color, and birth_year 
select(starwars, hair_color, eye_color, skin_color, birth_year) 
select(starwars, hair_color:birth_year) 
select(starwars, ends_with("color"), birth_year) 
select(starwars, ends_with("color"), ends_with("year")) 

``` 

## Adding Columns: 

`mutate()` takes your current columns, mutates them, and adds a new column to the end of the dataset. You can mutate either a single column or multiple columns. 

```{r} 

starwars_small = select(starwars, height, mass, birth_year) 
head(starwars_small) 

starwars_small = mutate(starwars_small, height_m = height/100) 

# you can mutate as many columns as you want in one function call 
mutate(starwars_small, bmi = mass/(height_m^2), new_birth_year = birth_year + 1)  

# if you only want to keep new columns, you use the transmute function:
transmute(starwars_small, height_m = height/100, bmi = mass/(height_m^2)) 

# using aggregate functions 
starwars_small = select(starwars, height, mass, birth_year) 
prop_mass = mutate(starwars_small, proportional_mass = mass/sum(mass, na.rm = T)) # na.rm - removes NA values 
arrange(prop_mass, desc(proportional_mass)) 

```


## Summarizing + Grouping Data Into One Value: 

`summarize()` or `summarise()` collapses an entire column of data to a single value 

```{r} 

mutate(starwars, mean_mass = mean(mass, na.rm = T)) 

# these two lines do the same thing: 
summarise(starwars, mean_mass = mean(mass, na.rm = T))
mean(starwars$mass, na.rm = T) 

## why you might want to use the summarize function instead of just calculating a number: 
# summarize() lets you calculate several "summary" values in one function call and it gives them to you in a table with 1 row 
# you can also group parts of the dataset: 
# n() returns the number of values in each group 
species_masses = summarise(group_by(starwars, species), mean_mass = mean(mass, na.rm = T), count = n()) 
# trying to see which species, on average, weighs the most: 
arrange(species_masses, desc(mean_mass)) 

## the pipe operator %>% 
# function(x, y) is the same as x %>% function(y) 
# everything below these comments does the same as the species_masses stuff above 

species_masses = starwars %>% 
  group_by(species) %>% 
  summarise(mean_mass = mean(mass, na.rm = T), count = n()) %>% 
  arrange(desc(mean_mass)) 
# when using dplyr functions, you generally always start with the name of the dataset 

``` 

## Sampling a Designated Number of Rows: 

`sample_n()` allows us to sample a random number of rows from our dataset. 

```{r} 

# every time this line is run, you get a different set of 10 random rows from the starwars dataset, but if you save it as an R object then it will always give you the same 10 rows every time you use it in your code 
sample_n(starwars, 10) 
starwars_10rows = sample_n(starwars, 10) 

# you can also use sample_frac() to get a certain percentage of the rows in a dataset (still randomly selected) 
nrow(starwars) # shows that the total number of rows in the starwars dataset is 87 
starwars_10percent = sample_frac(starwars, 0.1) 


starwars %>% 
  group_by(gender) %>% 
  sample_n(4) # pulls 4 rows from each gender; also lists NA values 

starwars %>% 
  group_by(species) %>% 
  sample_n(1) 

``` 

We can also take a "slice" of our dataset using `slice()` and its related set of functions. This retains the order of our rows in the dataset and is not random. 

```{r} 

slice(starwars, 50:60) # slice by index value (row number) 
slice_head(starwars, n = 10) #first 10 rows by index 
slice_tail(starwars, n = 10) # last 10 rows by index 
slice_min(starwars, order_by = mass, n = 10) # rows with lowest 10 masses 
slice_max(starwars, order_by = mass, n = 10) # rows with highest 10 masses  

``` 

`select()` can also be used to reorder columns: 

```{r} 

select(starwars, species, everything) # lists the species column first, then everything else comes after that 

select(starwars, species, hair_color, eye_color, height, everything()) 

``` 

```{r} 

filter(starwars, "A New Hope" %in% films) # shows the rows representing all the characters in the film "A New Hope" 

# this is different from filter(starwars, films == "A New Hope") because that would show the characters that were ONLY in the film "A New Hope", while filter(starwars, "A New Hope" %in% films) shows all the characters that were only in that film, along with the characters that were in that film as well as other star wars films 

```






